---
title: "Getting Around the tidyverse – Deconstructing Syntax and ggplot2"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r, message=F, warning=F, echo=F}
require(learnr)
require(tidyverse)
```

## "R"efresher

R is a scripting language commonly used by statisticians, data scientists, and people that need to deal with data. To deal with data, R has a lot of different ways that it can represent data, called data structures. Which ones are data structures in R?

```{r data-struc, echo = F}
question(
    "Which ones are data structures in R?",
    answer("data frame", T), answer("vector", T), answer("list", T), answer("factor", T), answer("matrix", T), answer("array", T),
    type = "multiple",
    correct = "Correct! All of them are data structure in R.",
    incorrect = "Incorrect! All of them are data structure in R."
)
```

### R as a Calculator

In R, we also have operators, so things like `+`, `-`, and `/` in math. One thing to keep in mind is that operations run in order of PEMDAS, so we have to be careful when doing math or other operations.

I'm trying to calculate $\frac{1+2+3}{2}$, but my code below isn't calculating the right answer, can you fix the code so that it calculates $3$?

```{r pemdas, exercise=TRUE}
# Fix this so that 3 is calculated
1 + 2 + 3 / 2
```

```{r pemdas-solution}
(1 + 2 + 3) / 2
```

### Assignment

R can also remember the values of things. Remember, we can do this by using `<-` or `=`.

```{r assignment-demo, exercise=TRUE}
x <- 1 + 1
x
y = 3 * 3
y
```

### ggplot review

R can also create visualizations, and a preferred way of creating visualization in R is by using the `ggplot2` package. All a package is in R (and in programming) is a collection of functions or other related things that allow you to use R at a much higher level (you don't need to know how the code is written for them), and for the most part, packages make your life a lot more easier since you can Google how to do something and there is most likely one that fulfills your needs.

Now, back to ggplot, remember that to make a plot

1. You have to first use `ggplot`.

2. You have to specify what kind of visualization ggplot should create. These would be like `geom_histogram`, `geom_point`, `geom_boxplot`, etc.

3. Make sure you specify the data (as a data frame) and aesthetic mapping in either `ggplot` or the function that specifies the visualization. And an aesthetic mapping is simply specifying the variables you want in a visualization.

As an example of what I mean by these points, take the following as example:

```{r ggplot2-demo, exercise=TRUE}
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
ggplot(data = iris) + geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width))
ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) + geom_point(data = iris)
ggplot() + geom_point(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width))
```

All of the plots are clearly the same, but the arguments were passed into different functions.

For more context on this behavior, consider the following plots (purely for demonstration; you'd never want to make a confusing plots like the last one):

```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Length)) + geom_point(mapping = aes(y = Sepal.Width))
ggplot(data = iris, mapping = aes(x = Sepal.Length)) + geom_point(mapping = aes(y = Petal.Width))
ggplot(data = iris, mapping = aes(x = Sepal.Length)) + geom_point(mapping = aes(y = Sepal.Width)) + geom_point(mapping = aes(y = Petal.Width))
```

What do you notice about the behavior of the aesthetic (or, in general, anything that can be put into `ggplot` or a function specifying a visualization type)?

## ggplot2 ideology

### Layers

All graphics made with ggplot are composed of layers. Of these layers, these are the most common layers that you'll encounter and questions to consider if you want to utilize them:

Layer       | Description and Guiding questions
----------  | -------------------------------------------
geometry    | Geometry to represent the data visually; what kind of plot do you want to create?
statistics  | An alternative to geometry
scales      | How numbers in data are converted to numbers on screen; are there any changes that you want to make to the tick marks on your axes or any changes relating to how the data is represented in the visualization?
labels      | Titles and axis labels; do you want to change the titles and axis labels so that it's more readable to a human?
guides      | Legend settings; do you want to edit your legend?
annotations | Additional geoms that are not mapped to data; do you want to add any text or other data that will enhance your visualization?
coordinates | Coordinate systems (Cartesian, logarithmic, polar)
facets      | Side-by-side panels; do you want to separate data in a different visualizations by a variable in your data frame?

We've already been utilizing the first type of layer in our past examples, so `geom_point`, `geom_histogram`, `geom_...` are all geometry layers (hence `geom`). Now, you may notice that overtime we create plots, we have to use the `+` operator to add geometry layers. It may seem counter intuitive, but think of it like this: you want to add layers to something, so you just simply add it to everything else. So, if you want to add a layer to a plot, you'll have to add it to the previous parts.

### The ggplot object

All graphics have one thing in common, they all need to have a `ggplot` object as the "base" of the visualization. So, you will always have to do something like `ggplot(...)` then add layers to the object so that something is produced. Without the object, no graphic will be outputted (you'll just get some text output detailing what you created).

### Try

Now that you know about layers, lets try to identify some layers that you'll need to create some visualizations:

Suppose that you are using the `cars` data set in R, and it has variables `speed` and `dist`; I run some code to produce the following:

```{r echo = F}
ggplot(cars, aes(speed, dist)) + geom_point() + labs(title = "Distance (miles) vs. Speed (mph)", x = "Speed (mph)", y = "Distance (miles)")
```

What types of layers did I use?

I used a data, aesthetic, geometry, and labels layer. Can you recreate the visualization that I made?

*Hint: I used `labs` as for my labels layer; it can take in `title`, `x`, and `y` arguments!*

```{r recreate-1, exercise = TRUE}
# replace the ... with your code! (you may not have to write code in all of them; you can just delete the ...)
ggplot(...) + geom_point(...) + labs(...)
```
